---
title: "Quantitative Analysis: Nutritional Benchmarking"
subtitle: "Comparing Co-designed Menus with Colombian Nutritional References"
author: "Arcila-Agudelo, A.M., Cardona-Trujillo, H., et al."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  cache: false
  eval: false
---

# Introduction

This analysis compares ideal menus co-designed in participatory workshops with Colombian nutritional references. The workshops were conducted in three regions (Medellín, Guarne, Santa Rosa de Osos) with three age groups (19-30, 31-50, 50+ years).

## Nutritional References

- **GABAS** (Guías Alimentarias Basadas en Alimentos): Colombian Food-Based Dietary Guidelines that recommend daily servings by food group
- **RIEN** (Recomendaciones de Ingesta de Energía y Nutrientes): Recommended intake of energy and nutrients for the Colombian population

## Analysis Flow

```{mermaid}
graph TD
    A[Bases de datos.xlsx] --> B[01_GenDataSet.R]
    B --> C[DS_Talleres_RIEN.rds]
    B --> D[DS_Talleres_GABAS.rds]
    C --> E[02_Gen_Graph_RIEN.R]
    D --> F[03_Gen_Graph_GABAS.R]
    C --> G[04_Gen_Graph_Nutrients.R]
    E --> H[RIEN Graphs]
    F --> I[GABAS Graphs]
    G --> J[Nutritional Heatmap]
```

# Setup and Dependencies

::: {.callout-note}
## Prerequisites
Before running this analysis, ensure you have:
- R (version 4.0 or higher)
- Required R packages: `tidyverse`, `readxl`, `writexl`, `ggplot2`, `reshape2`
- Input data file: `Bases de datos.xlsx`

Run `source("# Required libraries.r")` to install all dependencies.
:::

```{r setup}
#| label: setup
#| eval: true
#| code-summary: "Check environment"

cat("✓ R version:", R.version.string, "\n")
cat("✓ Working directory:", getwd(), "\n")

# Check for key input files
if (file.exists("Bases de datos.xlsx")) {
  cat("✓ Input data file found\n")
} else {
  cat("⚠ Warning: 'Bases de datos.xlsx' not found\n")
}

# Check for output datasets
if (file.exists("DS_Talleres_RIEN.rds")) {
  cat("✓ RIEN dataset found\n")
} else {
  cat("ℹ Note: Run 01_GenDataSet.R to generate RIEN dataset\n")
}

if (file.exists("DS_Talleres_GABAS.rds")) {
  cat("✓ GABAS dataset found\n")
} else {
  cat("ℹ Note: Run 01_GenDataSet.R to generate GABAS dataset\n")
}
```

# 1. Dataset Generation {#sec-dataset}

This step processes raw workshop data and generates structured datasets for comparison with nutritional references.

## 1.1 Load Workshop Data

```{r load-data}
#| label: load-data
#| code-summary: "Load raw workshop data"

# Read food weight data from workshops
ds <- readxl::read_xlsx("Bases de datos.xlsx", sheet = "Gramaje")

# Read GABAS equivalence table
eqv_gabas <- readxl::read_xlsx("Bases de datos.xlsx", 
                               sheet = "equiv_table_GABAS")

# Select and rename relevant columns
ds <- ds %>% 
  dplyr::select(
    ID, Region, Grupo, Plato, 
    `Grupo alimentario`, Alimento, `Porción final`,
    `Calorias porción elegida`, 
    `Proteina porción elegida`,
    `Carbohidratos por porción elegida`,
    `Grasas por porción elegida`,
    `Fibra por porción elegida`,
    `Calcio por porción elegida`,
    `Hierro por porción elegida`,
    `Sodio por porción elegida`
  )

# Rename for easier manipulation
names(ds) <- c(
  "ID", "Region", "Grupo", "Plato", "G_Alimento", "Alimento", "Porcion",
  "Calorias", "Proteina", "CHO", "Grasas", "Fibra", "Calcio",
  "Hierro", "Sodio"
)

# Merge with GABAS equivalence table
ds <- merge(ds, eqv_gabas)
ds <- ds %>% 
  select(ID, Region, Grupo, G_Alimento, G_Alimento_GABA, 
         Plato, Alimento:Sodio)

# Standardize region name
ds <- ds %>% 
  mutate(Region = ifelse(Region == "Medellín", "Medellin", Region))

cat("✓ Workshop data loaded and prepared\n")
cat("  - Total records:", nrow(ds), "\n")
cat("  - Regions:", paste(unique(ds$Region), collapse = ", "), "\n")
cat("  - Age groups:", paste(unique(ds$Grupo), collapse = ", "), "\n")
```

### Data Overview

```{r data-overview}
#| label: data-overview
#| tbl-cap: "Sample of workshop food selection data"

ds %>% 
  select(Region, Grupo, G_Alimento_GABA, Alimento, Porcion, Calorias) %>%
  head(10) %>%
  kable(digits = 2)
```

## 1.2 GABAS Analysis

Calculate total servings by region, GABAS food group, and age group.

```{r gabas-dataset}
#| label: gabas-dataset
#| code-summary: "Generate GABAS dataset"

# Calculate total servings by region, food group, and age group
ds_GABAS <- ds %>% 
  group_by(Region, G_Alimento_GABA, Grupo) %>% 
  summarise(Porcion = sum(Porcion, na.rm = TRUE), .groups = "drop")

# Load GABAS reference values
gabas_ref <- readxl::read_xlsx("Bases de datos.xlsx", 
                               sheet = "GABAS_Referencias")

# Merge with references
ds_GABAS <- merge(ds_GABAS, gabas_ref, 
                  by.x = c("G_Alimento_GABA", "Grupo"), 
                  by.y = c("G_Alimento_GABA", "Grupo"),
                  all.x = TRUE)

# Save dataset
saveRDS(ds_GABAS, "DS_Talleres_GABAS.rds")
write_xlsx(ds_GABAS, "DS_Talleres_GABAS.xlsx")

cat("✓ GABAS dataset generated\n")
cat("  - Food groups:", length(unique(ds_GABAS$G_Alimento_GABA)), "\n")
```

### GABAS Dataset Summary

```{r gabas-summary}
#| label: gabas-summary
#| tbl-cap: "GABAS servings by food group and region"

ds_GABAS %>%
  select(Region, Grupo, G_Alimento_GABA, Porcion, porcion_ref) %>%
  arrange(Region, Grupo, G_Alimento_GABA) %>%
  head(15) %>%
  kable(digits = 2, col.names = c("Region", "Age Group", "Food Group", 
                                    "Workshop Servings", "Reference Servings"))
```

## 1.3 RIEN Analysis

Calculate nutritional adequacy compared to RIEN recommendations.

```{r rien-dataset}
#| label: rien-dataset
#| code-summary: "Generate RIEN dataset"

# Aggregate nutrients by region and age group
ds_RIEN <- ds %>% 
  group_by(Region, Grupo) %>% 
  summarise(
    Calorias = sum(Calorias, na.rm = TRUE),
    Proteina = sum(Proteina, na.rm = TRUE),
    CHO = sum(CHO, na.rm = TRUE),
    Grasas = sum(Grasas, na.rm = TRUE),
    Fibra = sum(Fibra, na.rm = TRUE),
    Calcio = sum(Calcio, na.rm = TRUE),
    Hierro = sum(Hierro, na.rm = TRUE),
    Sodio = sum(Sodio, na.rm = TRUE),
    .groups = "drop"
  )

# Load RIEN reference values
rien_ref <- readxl::read_xlsx("Bases de datos.xlsx", sheet = "RIEN")

# Merge with references
ds_RIEN <- merge(ds_RIEN, rien_ref, by = "Grupo", all.x = TRUE)

# Save dataset
saveRDS(ds_RIEN, "DS_Talleres_RIEN.rds")
write_xlsx(ds_RIEN, "DS_Talleres_RIEN.xlsx")

cat("✓ RIEN dataset generated\n")
cat("  - Nutrients analyzed: 8 (Energy, Protein, CHO, Fat, Fiber, Ca, Fe, Na)\n")
```

### RIEN Dataset Summary

```{r rien-summary}
#| label: rien-summary
#| tbl-cap: "Nutritional adequacy: workshop menus vs RIEN recommendations"

ds_RIEN %>%
  select(Region, Grupo, Calorias, Calorias_ref, Proteina, Proteina_RDA) %>%
  mutate(
    Cal_Adequacy = round((Calorias / Calorias_ref) * 100, 1),
    Prot_Adequacy = round((Proteina / Proteina_RDA) * 100, 1)
  ) %>%
  select(Region, Grupo, Calorias, Calorias_ref, Cal_Adequacy, 
         Proteina, Proteina_RDA, Prot_Adequacy) %>%
  kable(col.names = c("Region", "Age Group", "Calories", "Ref", "% Adeq",
                       "Protein (g)", "Ref", "% Adeq"))
```

# 2. RIEN Visualization {#sec-rien}

Generate graphs comparing nutritional adequacy of co-designed menus with RIEN recommendations.

## 2.1 Graph Function

```{r rien-function}
#| label: rien-function
#| code-summary: "Define RIEN visualization function"

# Create combined label for visualization
ds_RIEN$x <- paste0(ds_RIEN$Region, " - ", ds_RIEN$Grupo)

# Function to create adequacy graphs
get_graph_r <- function(ds, val) {
  names(ds) <- c("x", "value1", "value2", "mymean")
  
  p <- ggplot(ds) +
    geom_point(aes(x = x, y = mymean), pch = 8, size = 2, shape = 18) +
    geom_segment(aes(x = x, xend = x, y = value1, yend = value2), 
                 color = "grey") +
    geom_point(aes(x = x, y = value1), 
               color = rgb(0.2, 0.7, 0.1, 0.5), size = 3) +
    geom_point(aes(x = x, y = value2), 
               color = rgb(0.7, 0.2, 0.1, 0.5), size = 3) +
    coord_flip() +
    theme_ipsum() +
    theme(legend.position = "none") +
    xlab("") +
    ylab(val)
  
  # Save graph
  ggsave(paste0("graphs/reg_", val, ".png"), plot = p, 
         width = 10, height = 6, dpi = 300)
  
  return(p)
}

cat("✓ RIEN visualization function defined\n")
```

## 2.2 Macronutrient Adequacy

### Energy (Calories)

```{r rien-calories}
#| label: rien-calories
#| fig-cap: "Energy adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Calorias_ref", "Calorias_ref", "Calorias")],
  "Calorias"
)
```

::: {.callout-note}
- **Green point**: Reference value (recommended energy intake)
- **Black diamond**: Average from co-designed menus
- **Gray line**: Difference between workshop value and reference
:::

### Protein

```{r rien-protein}
#| label: rien-protein
#| fig-cap: "Protein adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Proteina_RDA", "Proteina_EAR", "Proteina")],
  "Proteina"
)
```

### Carbohydrates

```{r rien-cho}
#| label: rien-cho
#| fig-cap: "Carbohydrate adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "CHO_EAR", "CHO_RDA", "CHO")],
  "CHO"
)
```

### Fats

```{r rien-fats}
#| label: rien-fats
#| fig-cap: "Fat adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Grasas_AI_min", "Grasas_AI_max", "Grasas")],
  "Grasas"
)
```

### Fiber

```{r rien-fiber}
#| label: rien-fiber
#| fig-cap: "Fiber adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Fribra_AI", "Fribra_AI", "Fibra")],
  "Fibra"
)
```

## 2.3 Micronutrient Adequacy

### Calcium

```{r rien-calcium}
#| label: rien-calcium
#| fig-cap: "Calcium adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Calcio_EAR", "Calcio_RDA", "Calcio")],
  "Calcio"
)
```

### Iron

```{r rien-iron}
#| label: rien-iron
#| fig-cap: "Iron adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Hierro_EAR", "Hierro_RDA", "Hierro")],
  "Hierro"
)
```

### Sodium

```{r rien-sodium}
#| label: rien-sodium
#| fig-cap: "Sodium adequacy by region and age group"

get_graph_r(
  ds_RIEN[, c("x", "Sodio_AI", "Sodio_AI", "Sodio")],
  "Sodio"
)
```

## 2.4 Overall Nutrient Comparison

### Macronutrients

```{r rien-macro-all}
#| label: rien-macro-all
#| fig-cap: "Macronutrient comparison across all groups"
#| fig-height: 8

macro_rien <- ds_RIEN %>% 
  select(Grupo, Proteina, CHO, Grasas, Proteina_EAR,
         Proteina_RDA, CHO_EAR, CHO_RDA, Grasas_AI_min, Grasas_AI_max)

# Reshape for visualization
macro_long <- macro_rien %>%
  select(Grupo, Proteina, CHO, Grasas) %>%
  pivot_longer(cols = -Grupo, names_to = "Nutrient", values_to = "Value")

ggplot(macro_long, aes(x = Grupo, y = Value, fill = Nutrient)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_ipsum() +
  labs(x = "Age Group", y = "Amount (g)", 
       title = "Macronutrient Content in Co-designed Menus") +
  theme(legend.position = "bottom")

ggsave("graphs/ALL_Macronutrientes.png", width = 10, height = 8, dpi = 300)
```

### Micronutrients

```{r rien-micro-all}
#| label: rien-micro-all
#| fig-cap: "Micronutrient comparison across all groups"
#| fig-height: 8

micro_rien <- ds_RIEN %>% 
  select(Grupo, Calcio, Hierro, Sodio)

# Reshape for visualization
micro_long <- micro_rien %>%
  pivot_longer(cols = -Grupo, names_to = "Nutrient", values_to = "Value")

ggplot(micro_long, aes(x = Grupo, y = Value, fill = Nutrient)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_ipsum() +
  labs(x = "Age Group", y = "Amount (mg)", 
       title = "Micronutrient Content in Co-designed Menus") +
  theme(legend.position = "bottom")

ggsave("graphs/ALL_Micronutrientes.png", width = 10, height = 8, dpi = 300)
```

# 3. GABAS Visualization {#sec-gabas}

Compare servings by food group with GABAS recommendations.

## 3.1 Graph Function

```{r gabas-function}
#| label: gabas-function
#| code-summary: "Define GABAS visualization function"

# Load GABAS dataset
ds_GABAS <- read_rds("DS_Talleres_GABAS.rds")

# Create combined label
ds_GABAS$x <- paste0(ds_GABAS$Region, " - ", ds_GABAS$G_Alimento_GABA)

# Function to create GABAS graphs
get_graph_gabas <- function(ds, titulo = "Porciones GABAS", filename = "GABAS") {
  
  p <- ggplot(ds, aes(x = x, y = Porcion)) +
    geom_point(aes(y = porcion_ref), 
               color = rgb(0.2, 0.7, 0.1, 0.7), 
               size = 4, shape = 18) +
    geom_segment(aes(x = x, xend = x, y = Porcion, yend = porcion_ref), 
                 color = "grey", linewidth = 1) +
    geom_point(color = rgb(0.7, 0.2, 0.1, 0.7), size = 3) +
    coord_flip() +
    theme_ipsum() +
    theme(legend.position = "none") +
    labs(title = titulo, x = "", y = "Servings")
  
  ggsave(paste0("graphs/", filename, ".png"), plot = p,
         width = 12, height = 8, dpi = 300)
  
  return(p)
}

cat("✓ GABAS visualization function defined\n")
```

## 3.2 GABAS Servings by Region

### Medellín

```{r gabas-medellin}
#| label: gabas-medellin
#| fig-cap: "GABAS servings comparison - Medellín"
#| fig-height: 8

ds_medellin <- ds_GABAS %>% filter(Region == "Medellin")
get_graph_gabas(ds_medellin, "GABAS - Medellín", "GABAS_Medellin")
```

### Guarne

```{r gabas-guarne}
#| label: gabas-guarne
#| fig-cap: "GABAS servings comparison - Guarne"
#| fig-height: 8

ds_guarne <- ds_GABAS %>% filter(Region == "Guarne")
get_graph_gabas(ds_guarne, "GABAS - Guarne", "GABAS_Guarne")
```

### Santa Rosa de Osos

```{r gabas-santarosa}
#| label: gabas-santarosa
#| fig-cap: "GABAS servings comparison - Santa Rosa de Osos"
#| fig-height: 8

ds_santarosa <- ds_GABAS %>% filter(Region == "Santa Rosa de Osos")
get_graph_gabas(ds_santarosa, "GABAS - Santa Rosa de Osos", "GABAS_SantaRosa")
```

## 3.3 GABAS Summary by Food Group

```{r gabas-summary-table}
#| label: gabas-summary-table
#| tbl-cap: "Average servings by food group across all regions"

ds_GABAS %>%
  group_by(G_Alimento_GABA, Grupo) %>%
  summarise(
    Workshop_Servings = mean(Porcion, na.rm = TRUE),
    Reference_Servings = mean(porcion_ref, na.rm = TRUE),
    Difference = Workshop_Servings - Reference_Servings,
    .groups = "drop"
  ) %>%
  arrange(G_Alimento_GABA, Grupo) %>%
  kable(digits = 2, col.names = c("Food Group", "Age Group", 
                                    "Workshop", "Reference", "Difference"))
```

# 4. Nutritional Heatmap {#sec-heatmap}

Visualize nutritional adequacy across all nutrients and groups.

```{r heatmap}
#| label: heatmap
#| fig-cap: "Nutritional adequacy heatmap"
#| fig-width: 12
#| fig-height: 8

# Calculate adequacy percentages
adequacy <- ds_RIEN %>%
  mutate(
    Cal_Adeq = (Calorias / Calorias_ref) * 100,
    Prot_Adeq = (Proteina / Proteina_RDA) * 100,
    CHO_Adeq = (CHO / CHO_RDA) * 100,
    Fat_Adeq = (Grasas / ((Grasas_AI_min + Grasas_AI_max) / 2)) * 100,
    Fiber_Adeq = (Fibra / Fribra_AI) * 100,
    Ca_Adeq = (Calcio / Calcio_RDA) * 100,
    Fe_Adeq = (Hierro / Hierro_RDA) * 100,
    Na_Adeq = (Sodio / Sodio_AI) * 100
  ) %>%
  select(Region, Grupo, ends_with("_Adeq"))

# Reshape for heatmap
adequacy_long <- adequacy %>%
  pivot_longer(cols = ends_with("_Adeq"), 
               names_to = "Nutrient", 
               values_to = "Adequacy") %>%
  mutate(
    Nutrient = str_remove(Nutrient, "_Adeq"),
    Group_Label = paste(Region, Grupo, sep = " - ")
  )

# Create heatmap
ggplot(adequacy_long, aes(x = Nutrient, y = Group_Label, fill = Adequacy)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Adequacy, 0)), color = "black", size = 3) +
  scale_fill_gradient2(
    low = "red", mid = "yellow", high = "green",
    midpoint = 100, limit = c(0, 200),
    name = "Adequacy (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(title = "Nutritional Adequacy Heatmap: Workshop Menus vs RIEN References")

ggsave("graphs/panel_b_heatmap.png", width = 12, height = 8, dpi = 300)
```

::: {.callout-tip}
## Interpretation Guide
- **Green cells (>100%)**: Nutrient exceeds recommendations
- **Yellow cells (~100%)**: Nutrient meets recommendations
- **Red cells (<100%)**: Nutrient below recommendations
:::

# 5. Key Findings

## 5.1 GABAS Compliance

```{r gabas-findings}
#| label: gabas-findings
#| code-summary: "Calculate GABAS compliance statistics"

gabas_compliance <- ds_GABAS %>%
  mutate(
    Difference = Porcion - porcion_ref,
    Percent_Diff = (Difference / porcion_ref) * 100,
    Status = case_when(
      abs(Percent_Diff) <= 10 ~ "Adequate",
      Percent_Diff > 10 ~ "Excessive",
      Percent_Diff < -10 ~ "Insufficient"
    )
  ) %>%
  count(Status) %>%
  mutate(Percentage = (n / sum(n)) * 100)

kable(gabas_compliance, digits = 1,
      col.names = c("Compliance Status", "Count", "Percentage (%)"),
      caption = "GABAS compliance summary")
```

## 5.2 RIEN Adequacy

```{r rien-findings}
#| label: rien-findings
#| code-summary: "Calculate RIEN adequacy statistics"

# Calculate mean adequacy across all groups
mean_adequacy <- adequacy %>%
  summarise(across(ends_with("_Adeq"), mean, na.rm = TRUE)) %>%
  pivot_longer(everything(), names_to = "Nutrient", values_to = "Mean_Adequacy") %>%
  mutate(
    Nutrient = str_remove(Nutrient, "_Adeq"),
    Status = case_when(
      Mean_Adequacy >= 90 & Mean_Adequacy <= 110 ~ "Adequate",
      Mean_Adequacy > 110 ~ "High",
      Mean_Adequacy < 90 ~ "Low"
    )
  )

kable(mean_adequacy, digits = 1,
      col.names = c("Nutrient", "Mean Adequacy (%)", "Status"),
      caption = "Mean nutritional adequacy across all groups")
```

## 5.3 Regional Differences

```{r regional-differences}
#| label: regional-differences
#| fig-cap: "Regional comparison of caloric adequacy"

ggplot(ds_RIEN, aes(x = Region, y = (Calorias / Calorias_ref) * 100, 
                     fill = Grupo)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
  theme_ipsum() +
  labs(x = "Region", y = "Caloric Adequacy (%)", 
       fill = "Age Group",
       title = "Caloric Adequacy by Region and Age Group") +
  theme(legend.position = "bottom")
```

# 6. Conclusions

## Main Findings

1. **GABAS Compliance**: The co-designed menus show varying compliance with GABAS recommendations across food groups.

2. **Nutritional Adequacy**: Mean adequacy varies by nutrient, with some nutrients meeting recommendations while others fall short.

3. **Regional Variations**: Significant differences exist between regions, reflecting local food preferences and availability.

4. **Age Group Considerations**: Nutritional adequacy differs by age group, with specific nutrients requiring attention in certain demographics.

## Recommendations

- Focus on increasing servings of food groups with low adequacy
- Address excessive servings where identified
- Consider regional preferences while promoting nutritional balance
- Tailor interventions to specific age group needs

# Session Information

```{r session-info}
#| label: session-info
#| code-summary: "R session information"

sessionInfo()
```

---

**Citation**: Arcila-Agudelo, A.M., Cardona-Trujillo, H., et al. (2026). Nutritional adequacy of co-designed food systems in rural Colombia. *Food Policy*, 199, 102756.

**Data Availability**: All data and code are available at [github.com/jcmunozmora/food-perception-rural-colombia](https://github.com/jcmunozmora/food-perception-rural-colombia)

**License**: CC-BY 4.0
